---
title: "Covid-19 Visualization"
author: "Jackson Wu"
output: html_document
---

# From epidemic to pandemic, visualizing the Covid-19 pandemic. Based and inspired by DataCamp projects.

## We will try to visualize the covid-19 pandemic during the beginning of the outbreak in 2020
```{r}
#necessary libraries
library(readr)
library(ggplot2)
library(dplyr)

#reading the dataset and assigning it as a variable

confirmed_cases_worldwide <- read.csv("confirmed_cases_worldwide.csv") # the file confirmed_cases_worldwide is a dataframe containing 2 column, date and cumulative sum of the confirmed covid 19 cases

confirmed_cases_worldwide$date <- as.Date(confirmed_cases_worldwide$date)
```


### How did the confirmed cases look around the world?


```{r}
# Draw a line plot of cumulative cases vs. date
ggplot(confirmed_cases_worldwide, aes(x = date, y = cum_cases)) +
  geom_line() +
  ylab("Cumulative confirmed cases")
```

We can see from the lineplot that the growth of confirmed cases was very extreme at the beginning of the 2020 outbreak, reaching a number of roughly 200 000 confirmed cases in just 2 months. It is also noticable that something happened during mid February that caused a weird jump.


Let now compare the world to China, the starting point of the pandemic.

```{r}
china_vs_world_cases <- read_csv("confirmed_cases_china_vs_world.csv")


china_vs_world <- ggplot(china_vs_world_cases) +
  geom_line(aes(date, cum_cases, color = is_china)) +
  ylab("Cumulative confirmed cases")

china_vs_world
```

We see that in China, cases jumped dramstically in mid-february, only for the line to flatten out, while rest of the is experiencing a exponential growth of cumulatitive cases.


Lets annotate and add WHO-events to get a more clear view of the there was a change as well why Chinas line began to flatten out.

```{r}
who_events <- tribble(
  ~ date, ~ event,
  "2020-01-30", "Global health\nemergency declared",
  "2020-03-11", "Pandemic\ndeclared",
  "2020-02-13", "China reporting\nchange"
) %>%
  mutate(date = as.Date(date))

# Using who_events, add vertical dashed lines with an xintercept at date
# and text at date, labeled by event, and at 100000 on the y-axis
china_vs_world +
  geom_vline(aes(xintercept = date), data = who_events, linetype = "dashed") +
  geom_text(aes(date, label = event), data = who_events, y = 1e5)
```

It looks like that the WHO declared a global health emergency as cases started increase heavily in early 2020 in China. We also see that Chinas case line in the plot jumped in mid-February due to country's changed policy in how cases were reported and contributed to the sudden jump. 

We can also see around mid-march where a global pandemic was declared as rest of the world was in the middle of a exponential growth of covid-cases.


lets add trend lines to get a deeper view on how the growth itself looked like.

```{r}
# Filter for China, from Feb 15
china_after_feb15 <- china_vs_world_cases %>%
  filter(is_china == "China", date >= "2020-02-15")

ggplot(china_after_feb15, aes(date, cum_cases)) +
  geom_line() +
  geom_smooth(method = "lm", se = FALSE) +
  ylab("Cumulative confirmed cases")
```

```{r}
# Filter confirmed_cases_china_vs_world for not China
not_china <- china_vs_world_cases %>%
  filter(is_china == "Not China")


not_china_trend_line <- ggplot(not_china, aes(date, cum_cases)) +
  geom_line() +
  geom_smooth(method = "lm", se = FALSE) +
  ylab("Cumulative confirmed cases")


not_china_trend_line
```

We can that growth rate in China slowed down after country changed the way it was reporting cases as the growth line was growing slower than the linear trend line.

The same could not be said about the rest of the world as the cases started to grow exponentially and outgrew the linear trendline.


### Who got hit hardest (other than china)

Lets investigate which countries got hit the hardest outside of China

```{r}
confirmed_cases_by_country <- read_csv("confirmed_cases_by_country.csv")

# Group by country, summarize to calculate total cases, find the top 7
top_countries_by_total_cases <- confirmed_cases_by_country %>%
  group_by(country) %>%
  summarize(total_cases = max(cum_cases)) %>%
  top_n(7, total_cases)


top_countries_by_total_cases
```
Lets try to visualize the data


```{r}
confirmed_cases_top7_outside_china <- read_csv("confirmed_cases_top7_outside_china.csv")


ggplot(confirmed_cases_top7_outside_china, aes(date, cum_cases, color = country)) +
  geom_line() +
  ylab("Cumulative confirmed cases")
```

We can see that Italy got hit the hardest outside of China in the beginning of the Covid-19 outbreak, for those that followed the news, this visualization might be no suprise, as Italy was the first country that reported rapid growth of Covid-19 cases in Europe. 





